Subgroup detection in linear growth curve models with generalized linear mixed model (GLMM) trees

### 2023-09-07 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
Paper updates
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3275 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 36a1766808faf6884af890d2c27118e77548feff

\documentclass[doc,floatsintext,natbib]{apa7}
%\usepackage[man,floatsintext]{apa6}
%\usepackage[doc,floatsintext]{apa6}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{doi}
\usepackage{booktabs}
%\usepackage[dvipsnames]{xcolor}
\usepackage{placeins}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{setspace}
\usepackage{hyperref}

\renewcommand{\doitext}{} % prevents superfluous addition of "doi:" in references
\renewcommand{\thefootnote}{\fnsymbol{footnote}} % uses symbols instead of numbers for footnotes

@@

\title{Subgroup detection in linear growth curve models with GLMM trees}
\shorttitle{Recursive partitioning of linear growth curve models}
\authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
\authorsaffiliations{{$^1$Leiden University}, {$^2$Universit\"at Innsbruck}}

\abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time. Heterogeneity between subjects is common in such models, and researchers are typically interested in explaining or predicting this heterogeneity. We show how generalized linear mixed effects model (GLMM) trees can be used to identify subgroups with differently shaped trajectories in linear growth curve models. Originally developed for clustered cross-sectional data, GLMM trees are extended here to longitudinal data. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case. We evaluate GLMM trees' performance in partitioning growth curve models through a simulation study and through application to children's development of scholastic abilities. We compare the performance of GLMM trees with alternative methods: structural equation model (SEM) trees and LongCART. GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART. In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.\\}

\begin{document}
\SweaveOpts{concordance=TRUE}
\maketitle

\section{Introduction}

\label{sec:Introduction}

@@

Development over time is of prime interest in psychological research. For example, in educational studies researchers may want to model student's academic development over time; in clinical studies researchers may want to model patients' symptoms over time. Mixed-effects or latent-variable models can be used to model such trajectories and allow for explaining heterogeneity with covariates of a-priori known relevance \citep[e.g., ][]{NeisyMatt18}. However, when the covariates that can explain heterogeneity are not known in advance, methods for identifying them are needed.

For example, Figure~\ref{fig:global_trajectories} shows trajectories of science knowledge and skills from 250 children who were assessed at three timepoints throughout grades 3 through 8\footnote{Further details on the source of these data and scaling of the response variable are provided in Study~III.}. The gray lines depict individual trajectories, while the red line depicts the estimated average trajectory. Although the red line describes the average trajectory well, the gray lines indicate substantial intra-individual variability in average levels and growth over time. A common research aim might be to identify which covariates can explain or predict this heterogeneity.

\begin{figure}%
\caption{Growth curves of science ability.}
\begin{subfigure}{0.65\textwidth}%
<<fig=TRUE, echo=FALSE, results=hide, fig.width=7, fig.height=7>>=
y <- t(sapply(unique(data$CHILDID), function(x) data$score[data$CHILDID == x]))
x <- t(sapply(unique(data$CHILDID), function(x) data$months[data$CHILDID == x]))
plot(data$months, data$score, type = "n", xlab = "Months after baseline", 
     ylab = "Science ability", xaxt = "n", cex.lab = 1, cex.axis=1,
     xlim = c(0.5, 57^(2/3)), main = "Full sample (N = 250)")
axis(1, at=c(0, 12, 36, 60)^(2/3), labels=c(0, 12, 36, 60), cex.axis = .8)
sapply(1:nrow(x), function(row) lines(x[row, ], y[row, ], col = gray(0.2, alpha = 1/6), lwd = 2))
lines(x = c(min(data$months), max(data$months)), 
     y = c(fixef(LMM)[1], fixef(LMM)[1] + max(data$months)*fixef(LMM)[2]),
     lwd = 3, col = 2)
@
\\\footnotesize{\textit{Note.} Gray lines depict observed individual trajectories. The red line depicts the average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals.}
\end{subfigure}%
\label{fig:global_trajectories}
\end{figure}%

\begin{figure}%
\caption{Partitioned growth curves of science abilities.}
\begin{subfigure}{0.8\textwidth}%
<<fig=TRUE, height=9, width=12, echo=FALSE>>=
LMM_form <- score ~ months | CHILDID | GENDER + RACE + SES + GMOTOR + FMOTOR + 
  INTERN + EXTERN + INTERP + CONTRO + FIRKDG + AGEBASELINE
testdat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
testdat$CHILDID <- factor(testdat$CHILDID)
testdat$GENDER <- factor(testdat$GENDER)
testdat$asmtmm <- factor(testdat$asmtmm)
## Remove obs with levels of race in test that are not in train
levs <- unique(testdat$RACE) %in% unique(data$RACE)
if (!all(levs)) {
  testdat <- testdat[-which(!testdat$RACE %in% unique(testdat$RACE)[levs]), ]
}
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
plot(lmm_tree_c, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = 2, tp_args = list(xscale = c(0, 60^(2/3)), 
                                        xaxis.at = c(0, 12, 36, 60)^(2/3),
                                        xaxis.labs = c(0, 12, 36, 60)),
     gp = gpar(cex = 1.5))
icc_c <- as.data.frame(VarCorr(lmm_tree_c))
lmm_c_preds <- predict(lmm_tree_c, newdata = testdat, re.form=NA)
lmm_preds <- predict(LMM, newdata = testdat, re.form=NA)
@
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals.}
\end{subfigure}%
\label{fig:LMM_tree_c}
\end{figure}%

<<echo=FALSE, results=tex, fig.pos='b'>>=
library("kableExtra")
coefs <- coef(lmm_tree_c)
coefs <- cbind(Node = as.numeric(rownames(coefs)), coefs)
colnames(coefs) <- c("Node", "Intercept", "Slope")
kable_styling(kable(coefs, format = "latex", booktabs = TRUE, label = "local_coefs", 
                             align = "lccc", escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
        caption = "Fixed-effects coefficient estimates from the terminal nodes of Figure 2.", 
        row.names = FALSE, digits = 3L),
  font_size = 11, full_width=FALSE)
@

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Figure~\ref{fig:LMM_tree_c} shows an example tree, which identified socio-economic status (SES) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of these children, assessed at baseline. Four subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope (see Table~\ref{tab:local_coefs}). For children in the group with lower levels of SES, the intercept is also lowest, indicating lower science ability. For intermediate levels of SES (between $-0.41$ and $1.23$), there is also an effect of internalizing problems, where children with higher internalizing problems have a lower intercept. Highest science ability intercepts are observed in the group with highest SES. It appears that nodes with higher intercepts also tend to have higher slopes, indicating that children with relatively higher ability also show larger ability gains over time.

The tree in Figure~\ref{fig:LMM_tree_c} has been estimated with generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}. GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are for example nested in treatment centers, classrooms and/or geographical areas. In the current paper we aim to extend GLMM trees, so that they can be applied to partitioning LGCMs. The general idea of GLMM trees is appealing for subgroup detection in almost any type of mixed-effects model. Compared to clustered cross-sectional data, however, longitudinal data may require a different estimation approach: The variance of random effects tends to be higher with longitudinal data, and predictors of interest tend to be measured at higher levels (e.g., time-invariant covariates). In this paper, we propose and test two extensions of GLMM trees that account for these characteristics. We focus on the specific use case of partitioning LGCMs, but the extensions are critical for a wider range of settings where covariates are measured at higher levels and/or where the random effects have substantial variance.

There are several alternative RPMs that can be used to partition linear growth-curve models (LGCMs): GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19}. Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods have a specific aim of detecting subgroups for which the time-by-treatment interaction differs, so are not general methods for partitioning growth curve models.}.

@@

Other methods for partitioning LGCMs take a fully local estimation approach: Within every node or subgroup defined by the terminal nodes, a full parametric model is estimated based on the observations in that subgroup only. This fully local estimation approach provides more flexibility but also comes at a higher computational burden and increased model complexity. GLMM trees require estimation of a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the results. Furthermore, the fully local estimation approach requires possible partitioning variables to be measured at the highest level of the multilevel structure, while the local-global estimation approach allows for partitioning variables to be measured at any level.

The computational advantage of GLMM trees is strongest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees. These methods employ an exhaustive split detection procedure, where for every possible split point in the current node, the full parametric model needs to be re-estimated in the two resulting nodes. To choose the optimal split, the splitting criterion (such as a $p$-value from a likelihood-ratio test) is derived from these two models. Not only does this bring a heavy computational load, it also introduces a selection bias towards covariates with a larger number of possible cutpoints \citep{ShihyTsai04, Shih04}. LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not require model refitting for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This reduces computational load, while it also allows for separating variable and cutpoint selection, thus preventing selection bias. The GLMM tree algorithm also employs a two-step approach to split selection and thereby shares these advantages.

Given this lower model complexity, lower computational burden and lack of variable selection bias, GLMM trees might be particularly useful for subgroup detection in LGCMs. In the remainder of this section, we explain how GLMM trees are estimated and propose adjustments for partitioning longitudinal trajectories. Next, we assess the performance of the proposed adjustments to GLMM trees in partitioning LGCMs: In Study~I, we assess performance of the proposed adaptations to GLMM trees in simulated datasets. In Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART. In the Study~III, we compare the performance of the proposed adjustments of GLMM trees in existing datasets on children's development of reading, math and science abilities. In the Discussion, we summarize our findings and discuss implications.

@@

\section{Estimation of GLMM trees}
\label{sec:estimation}

@@

The following two subsections describes an alternative initialization strategy for Step~0 and an alternative approach for fitting the GLM tree from Step~1. Each subsection first reviews the well-established methods and then proceeds to discuss modifications that may improve performance when partitioning longitudinal data.

@@

Previous publications on recursive partitioning of models with mixed effects find that initializing the random-effect estimates with zero yields accurate estimates of the subgroup memberships and thus of the final models  \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} specifically assessed the impact of different initialization values and found only minor differences that decreased further with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data. With longitudinal data, however, random effects tend to be more pronounced: Repeated measures on the same subjects tend to be correlated more strongly than observations nested within the same unit in cross-sectional data. If the random effects are sizable, the initial assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations. Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery. Specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.

@@

\begin{figure}%
\caption{GLMM tree estimated by initializing with zero random effects.}
<<fig=TRUE, height=7, width=12, echo=FALSE, out.width=12>>=
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
plot(lmm_tree, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1))
icc <- as.data.frame(VarCorr(lmm_tree))
lmm_preds <- predict(lmm_tree, newdata = testdat, re.form=NA)
@
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, the $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
\label{fig:LMM_tree}
\end{figure}%

\begin{figure}%
\caption{GLMM tree initialized by estimating the random effects.}
<<fig=TRUE, height=9, width=15, echo=FALSE>>=
lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
icc_r <- as.data.frame(VarCorr(lmm_tree_r))
lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
@
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
\label{fig:LMM_tree_r}
\end{figure}%

To illustrate how initialization may make a difference we applied both approaches to the dataset from Figures~\ref{fig:global_trajectories}~and~\ref{fig:LMM_tree_c}. Initializing estimation assuming zero random effects yielded the tree in Figure~\ref{fig:LMM_tree}, with \Sexpr{width(lmm_tree$tree)} subgroups and an estimated random-intercept variance of \Sexpr{formatC(data.frame(VarCorr(lmm_tree$lmer))$vcov[1], digits = 2)}. Additional splits were implemented based on gender, race, internalizing problems and fine motor skills. Considering the relatively large number of subgroups and relatively small sample sizes in some of the terminal nodes, this tree may not generalize well to other samples. Initializing with estimation of the random effects yielded the tree depicted in Figure~\ref{fig:LMM_tree_r}, with \Sexpr{width(lmm_tree_r$tree)} subgroups and an estimated random-intercept variance of \Sexpr{formatC(data.frame(VarCorr(lmm_tree_r$lmer))$vcov[1], digits = 2)}. Compared to the tree in Figure~\ref{fig:LMM_tree_c}, an additional split was implemented based on gross motor skills (GMOTOR): children with further developed motor skills at baseline have a higher intercept.

\subsection{Partitioning}

>>>>

\documentclass[doc,floatsintext,natbib]{apa7}
%\usepackage[man,floatsintext]{apa6}
%\usepackage[doc,floatsintext]{apa6}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{doi}
\usepackage{booktabs}
%\usepackage[dvipsnames]{xcolor}
\usepackage{placeins}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{setspace}
\usepackage{hyperref}
\renewcommand{\doitext}{} % prevents superfluous addition of "doi:" in references
\renewcommand{\thefootnote}{\fnsymbol{footnote}} % uses symbols instead of numbers for footnotes

@@

\title{Subgroup detection in linear growth curve models with generalized linear mixed model (GLMM) trees}
\shorttitle{Recursive partitioning of linear growth curve models}
\authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
\authorsaffiliations{{$^1$Leiden University}, {$^2$Universit\"at Innsbruck}}

\abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time. Heterogeneity between subjects is common in such models, and researchers are typically interested in explaining or predicting this heterogeneity. We show how generalized linear mixed effects model (GLMM) trees can be used to identify subgroups with differently shaped trajectories in linear growth curve models. Originally developed for clustered cross-sectional data, GLMM trees are extended here to longitudinal data. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case. In simulated and real-world data, we assess the performance of the extensions and compare against other partitioning methods for growth curve models. Extended GLMM trees perform more accurately than the original algorithm and LongCART, and similarly accurate as structural equation model (SEM) trees. In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and are much faster to compute.\\}

\begin{document}
\SweaveOpts{concordance=TRUE}
\maketitle
\setlength{\tabcolsep}{3pt}

\section{Introduction}
\label{sec:Introduction}

@@

Development over time is of prime interest in psychological research. For example, in educational studies researchers may want to model student's academic development over time; in clinical studies researchers may want to model patients' symptoms over time. Mixed-effects or latent-variable models can be used to model such trajectories and allow for explaining heterogeneity with covariates of a-priori known relevance \citep[e.g., ][]{NeisyMatt18}. However, when these covariates are not known in advance, methods for identifying them are needed.

As an example, trajectories of science knowledge and skills from a sample of 250 children are depicted in Figure~\ref{fig:global_trajectories}. The children were assessed at three timepoints across grades 3 through 8\footnote{Further details on the source of these data are provided in Study~III.}. The red line depicts the estimated average trajectory, while the gray lines depict individual trajectories. The gray lines reveal substantial variability between the children, both in initial levels and growth over time. An obvious research aim would be to identify covariates that can explain or predict this heterogeneity.

\begin{figure}[b]%
\caption{Growth curves of science ability.}
\begin{subfigure}{0.65\textwidth}%
<<fig=TRUE, echo=FALSE, results=hide, fig.width=7, fig.height=7>>=
y <- t(sapply(unique(data$CHILDID), function(x) data$score[data$CHILDID == x]))
x <- t(sapply(unique(data$CHILDID), function(x) data$months[data$CHILDID == x]))
plot(data$months, data$score, type = "n", xlab = "Months after baseline", 
     ylab = "Science ability", xaxt = "n", cex.lab = 1, cex.axis=1,
     xlim = c(0.5, 57^(2/3)), main = "Full sample (N = 250)")
axis(1, at=c(0, 12, 36, 60)^(2/3), labels=c(0, 12, 36, 60), cex.axis = .8)
sapply(1:nrow(x), function(row) lines(x[row, ], y[row, ], col = gray(0.2, alpha = 1/6), lwd = 2))
lines(x = c(min(data$months), max(data$months)), 
     y = c(fixef(LMM)[1], fixef(LMM)[1] + max(data$months)*fixef(LMM)[2]),
     lwd = 3, col = 2)
@
\\\footnotesize{\textit{Note.} Gray lines depict observed individual trajectories, red line depicts average growth curve as estimated with a linear mixed-effect model, comprising a fixed effect of time and a random intercept with respect to individuals. The $x$-axis is not linear in the number of months, because time was scaled as $(\text{\# of months})^\frac{2}{3}$ in order to obtain approximately linear trajectories.}
\end{subfigure}%
\label{fig:global_trajectories}
\end{figure}%

%\FloatBarrier
\subsection{Recursive Partitioning Methods for Growth Curve Models}

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Figure~\ref{fig:LMM_tree_r} shows an example tree, which identified socio-economic status (SES), gross motor skills (GMOTOR) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of the children, assessed at baseline. Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope %(presented in Table~\ref{tab:local_coefs}). 
Groups of children with lower (higher) SES also have a lower (higher) intercept, indicating lower (higher) average science ability. For children with lower SES (node~2), there is also an effect of gross motor skills: The group with lower motor skills has a lower intercept. For children with intermediate levels of SES (node~6), there is also an effect of internalizing problems: The group with lower internalizing problems has a higher intercept. Also, the two groups (or nodes) with higher intercepts have higher slopes, indicating that children with relatively higher ability may also show larger gains in ability over time.

\begin{figure}%
\caption{Partitioned growth curves of science ability.}
\begin{subfigure}{0.9\textwidth}%
<<fig=TRUE, height=9, width=13.5, echo=FALSE>>=
LMM_form <- score ~ months | CHILDID | GENDER + RACE + SES + GMOTOR + FMOTOR + 
  INTERN + EXTERN + INTERP + CONTRO + FIRKDG + AGEBASELINE
testdat <- sciedata[!sciedata$CHILDID %in% bag_ids[ , 39], ]
testdat$CHILDID <- factor(testdat$CHILDID)
testdat$GENDER <- factor(testdat$GENDER)
testdat$asmtmm <- factor(testdat$asmtmm)
## Remove obs with levels of race in test that are not in train
levs <- unique(testdat$RACE) %in% unique(data$RACE)
if (!all(levs)) {
  testdat <- testdat[-which(!testdat$RACE %in% unique(testdat$RACE)[levs]), ]
}

## Fit models
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
lmm_tree_c <- lmertree(LMM_form, data = data, cluster = CHILDID, maxdepth = 6)
lmm_tree_r <- lmertree(LMM_form, data = data, maxdepth = 6, ranefstart = TRUE)
icc_r <- as.data.frame(VarCorr(lmm_tree_r))
lmm_r_preds <- predict(lmm_tree_r, newdata = testdat, re.form=NA)
icc_c <- as.data.frame(VarCorr(lmm_tree_c))
lmm_c_preds <- predict(lmm_tree_c, newdata = testdat, re.form=NA)
lmm_preds <- predict(LMM, newdata = testdat, re.form=NA)

## Plot tree
plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
@
\label{subfig:tree}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals.}
\end{subfigure}%
~
\begin{subfigure}{0.25\textwidth}%
\vspace{-2.95cm}
\hspace{-2.5cm}
\scriptsize
<<echo=FALSE, results=tex, fig.pos='b'>>=
library("kableExtra")
coefs <- coef(lmm_tree_r)
coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
colnames(coefs) <- c("Node", "Intercept", "Slope")
coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
                                                       coefs$Intercept[!grepl("-", coefs$Intercept)])
kable(coefs, format = "latex", linesep = "", # linesep command suppressess addlinesep every 5 rows
        row.names = FALSE, digits = 3L,escape=FALSE, align = c("ccc"), booktabs = TRUE)
@
\label{subfig:coefs}
\end{subfigure}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories, red lines depict average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals. The table presents numerical estimates of fixed intercepts and slopes.}
\label{fig:LMM_tree_r}
\end{figure}%

<<echo=FALSE, results=tex, fig.pos='b'>>=
# library("kableExtra")
# coefs <- coef(lmm_tree_r)
# coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
# colnames(coefs) <- c("Node", "Intercept", "Slope")
# coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
# coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
# coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
#                                                        coefs$Intercept[!grepl("-", coefs$Intercept)])
# kable_styling(kable(coefs, format = "latex", booktabs = TRUE, label = "local_coefs",
#                              align = "lccc", escape=FALSE, linesep = "", # linesep command suppressess addlinesep every 5 rows
#         caption = "Fixed-effects coefficient estimates from the terminal nodes of Figure 2.",
#         row.names = FALSE, digits = 3L),
#   font_size = 11, full_width=FALSE)
@

The tree in Figure~\ref{fig:LMM_tree_r} has been estimated with generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}. GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are nested in treatment centers, classrooms and/or geographical areas, for example. In the current paper we extend GLMM trees, so that they can be applied to partitioning LGCMs. The general idea of GLMM trees is appealing for subgroup detection in almost any type of mixed-effects model. Compared to clustered cross-sectional data, however, longitudinal data may require a different estimation approach: The variance of random effects tends to be higher with longitudinal data, and the predictors of interest tend to be measured at higher levels (e.g., time-invariant covariates). In this paper, we propose and test two extensions of GLMM trees that account for these characteristics. We focus on the specific use case of partitioning LGCMs, but the extensions are critical for a wider range of settings where covariates are measured at higher levels and/or where the random effects have substantial variance.

There are several alternative RPMs that can be used to partition linear growth-curve models (LGCMs): GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19}. Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods specifically target subgroups with different time-by-treatment interactions, so are not generally applicable for partitioning growth curve models.}.

@@

Other methods for partitioning LGCMs take a fully local estimation approach: Within every node or subgroup defined by the terminal nodes, a full parametric model is estimated based on the observations in that subgroup only. This fully local estimation approach provides more flexibility, but yields higher computational burden and model complexity. In contrast, GLMM trees estimate a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the results. Furthermore, the fully local estimation requires possible partitioning variables to be measured at the highest level of nesting, while GLMM trees' local-global estimation approach allows partitioning variables to be measured at any level.

The computational advantage of GLMM trees is strongest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees. These methods employ an exhaustive split detection procedure, where for every possible split point in the current node, the full parametric model needs to be re-estimated in the two resulting nodes. To choose the optimal split, the splitting criterion (such as a $p$-value from a likelihood-ratio test) is derived from these two models. Not only does this bring a heavy computational load, it also introduces a selection bias towards covariates with a larger number of possible cutpoints \citep{ShihyTsai04, Shih04}. LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not require model refitting for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This reduces computational load, while it also allows for separating variable and cutpoint selection, thus preventing selection bias. The GLMM tree algorithm shares these advantages, because it also employs a two-step approach to split selection.

Given their unbiased variable selection, lower model complexity and computational burden, GLMM trees might be particularly useful for subgroup detection in LGCMs. The next section explains how GLMM trees are estimated and propose adjustments for partitioning longitudinal trajectories. Next, the performance of the proposed adjustments is evaluated: In Study~I, we assess performance in simulated datasets, in Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART. In Study~III, we assess performance of the proposed adjustments in existing datasets on children's development of reading, math and science abilities. In the Discussion, we summarize our findings and discuss implications.

@@

\FloatBarrier
\section{Estimation of GLMM trees and Adaptations for Longitudinal Data}
\label{sec:estimation}

@@

The following two subsections describe alternative approaches for the initialization in Step~0 and for fitting the GLM tree in Step~1. Each subsection first reviews the well-established methods and then proceeds to discuss modifications that may improve performance when partitioning longitudinal data.

@@

Previous publications on mixed-effects recursive partitioning find that initializing the random-effect estimates with zero yields accurate estimates of subgroup memberships and the final models \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} assessed the impact of different initialization values and found only minor differences that decreased with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data. With longitudinal data, however, random effects tend to be more pronounced: Repeated measures on the same subjects tend to be correlated more strongly than observations nested within the same unit in cross-sectional data. If random effects are sizable, the initial assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations. Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery. Specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.

@@

\begin{figure}%
\caption{GLMM tree estimated by initializing with zero random effects.}
<<fig=TRUE, height=8, width=14, echo=FALSE>>=
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
plot(lmm_tree, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1))
icc <- as.data.frame(VarCorr(lmm_tree))
lmm_preds <- predict(lmm_tree, newdata = testdat, re.form=NA)
@
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, the $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
\label{fig:LMM_tree}
\end{figure}%

To illustrate, we applied both initialization approaches to the dataset from Figure~\ref{fig:global_trajectories}. In fact, the tree presented in Figure~\ref{fig:LMM_tree_r} was estimated by initializing with the random effects. Initializing estimation assuming zero random effects resulted in the tree in Figure~\ref{fig:LMM_tree}. The split based on gross motor skills in Figure~\ref{fig:LMM_tree_r} was not implemented in  Figure~\ref{fig:LMM_tree}, while additional splits were implemented based on gender, race, internalizing problems and fine motor skills. Considering the relatively large number of subgroups in  Figure~\ref{fig:LMM_tree_r}, some with relatively small sample sizes, this tree may overfit and not generalize well to other samples.

\FloatBarrier
\subsection{Partitioning}

\begin{figure}%
\caption{GLMM tree estimated with cluster-level parameter stability tests.}
<<fig=TRUE, height=8, width=12, echo=FALSE>>=
plot(lmm_tree_c, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = 2, tp_args = list(xscale = c(0, 60^(2/3)), 
                                        xaxis.at = c(0, 12, 36, 60)^(2/3),
                                        xaxis.labs = c(0, 12, 36, 60)),
     gp = gpar(cex = 1.5))
@
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
\label{fig:LMM_tree_c}
\end{figure}%
